Boosting, Sparsity, and Conditional Gradient: A Tutorial

### 2014-02-18 arXiv/introduction.tex
added a bit of the intro

Many algorithms, such as SMC \citep{smc} and MCMC \citep{mcmc}, approximate a 
distribution with samples. Most such algorithms require access to a full posterior and 
may be prone to becoming stuck in local modes. In contrast, the herding algorithm introduced in 
\citet{Welling:2009a} generates samples whose moments are guaranteed to converge 
to any given collection of feasible moments. Moreover, \emph{only} the moments 
are required; a full posterior is not needed.

Since its introduction, herding has been shown to have many desirable properties.
For instance, \citet{Chen:2010a} showed that the moments converge at a rate of $O(1/T^{2})$ for 
$T$ samples, which is even faster than the $O(1/T)$ rate that would be obtained 
by sampling independently from the exact posterior. Furthermore, \citet{Bach:2012a} show that herding is 
equivalent to conditional gradient, and \citet{Huszar:2012} show that 
herding is equivalent to Bayesian quadrature. Herding was originally conceived of as a method 
for inference in Markov random fields, and this line of work has been further explored in 
\citet{Welling:2009a}, \citet{Gelfand:2010}, and \citet{Bornn:2013}.

In order to define herding, let $\sX$ be an observation space, $\phi : \sX \to \sF$ a 
feature map to an inner product feature space, and $\bar\phi = \bE_{x \sim p}[\phi(x)]$ the empirical 
moments of some distribution. Then the herding algorithm is the following simple procedure to 
generate pseudosamples from $\sX$:
\[
x_{t} &\in \argmin_{x \in \sX} \ip{\phi(x)}{y_{t-1}} \\
y_{t} &= y_{t-1}  + \bar \phi - \phi(x_{t}).
\]
In this paper we generalize the herding algorithm to minimize $\psi(x) + R^{*}(x)$ via the following set of updates, which 
require only the convexity of $\psi$ and $R$:
\[
x_t &\in \argmin_{x} \psi(x) + \ip{x}{y_{t-1}} \\
y_{t} &\in \argmax_{y} \ip{\bar{x}_t}{y} - R(y),
\]
where $\bar{x}_t$ is a certain weighted average of $x_1,\ldots,x_t$. Our 
algorithm is described in more detail in Section~\ref{sec:algorithm}. This 
generalization has several advantages:
\begin{itemize}
\item It allows us to cast herding in the framework of mirror descent \cite{Beck:2003}, 
      which lets us better understand its convergence properties.
\item It allows us to obtain generalizations of herding for infinite 
      dimensional spaces that apply in cases where standard herding does not.
\item It makes the known connections between herding, subgradient, and 
      conditional gradient methods more clear. 
\end{itemize}
In particular, while it was known that herding was a form of conditional 
gradient descent \cite{Bach:2012a} and hence (via the relationship in \citet{Bach:2012b}) also 
a form of subgradient descent, our framework exposes all three algorithms 
as special cases. A full table outlining the relationships for various choices of 
$\psi$ and $R$ is given in Table~\ref{tab:connections}.

In summary, we introduce a novel gradient descent algorithm to investigate herding and its generalizations, and make the following contributions:
\begin{itemize}
\item In Section \ref{sec:algorithm} we introduce a pair of convex optimization algorithms, which we call primal and dual boosted mirror descent (\primal and \dual). We show how \primal and \dual relate to conditional gradient (\cgd) and mirror descent (\md). 
\item We then use \dual in Section \ref{sec:herding} to define two classes of generalized herding algorithms. Connections with the ``learning'' interpretation of herding are made by showing how to obtain herding as the limit of an entropically normalized case of \dual. 
\item Section \ref{sec:proofs} presents convergence rates for \primal and \dual. As special cases, we recover a $O(\log T/ T)$ convergence rate for classical herding and a $O(1/T)$ rate for weighted herding. Both of these results apply to the infinite dimensional case. 
\item In Section \ref{sec:lower-bounds} we prove a matching lower bound to the upper bound given in Section \ref{sec:proofs}, showing that any ``sparse'' herding-like method cannot converge faster than $O(1/T)$ in the the infinite dimensional case.% and that herding converges at the slower $O(\log T / T)$ rate.
%\item Finally, in section \ref{sec:chen-proofs} we generalize the $O(1/T^{2})$ convergence proof of \citet{Chen:2010a} to a class of herding-like algorithms derived from \bmd. 
\end{itemize}

>>>>

The purpose of this paper is to serve as a tutorial on sparsification. While 
most of the results presented are known, they are often distributed across the 
theory, machine learning, and optimization literatures, and we hope to bring 
them all into one place, as well as to provide a unifying perspective. To start, 
we define a generalized notion of sparsity that encompasses both sparsity and 
low-rank conditions as special cases. Throughout, we will let $S$ denote a 
compact subset of $\bR^n$ and let $\Conv(S)$ denote the convex hull of $S$ 
(which will always be a closed set due to compactness of $S$). Two particularly 
important examples are:
\begin{example}
\label{ex:simplex}
Let $S$ consist of the $n$ standard unit vectors $e_1, \ldots, e_n$. Then 
$\Conv(S)$ is the $n$-dimensional simplex $\Delta_n$.
\end{example}

\begin{example}
\label{ex:sdp}
Let $S = \{vv^{\top} \mid \|v\|_2 = 1\}$. An alternative characterization 
of $S$ is all rank-$1$, positive semidefinite matrices with trace $1$. The 
convex hull of $S$ is $\Conv(S) = \{V \mid V \succeq 0, \Tr(V) = 1\}$.
\end{example}

\begin{definition}
Given a set $S \subseteq \bR^n$, a \emph{$(S, k)$-sparse} point is a point 
$x \in \Conv(S)$ that is a convex combination of at most $k$ extreme 
points of $\Conv(S)$.
\end{definition}

Recall that an extreme point is a point that cannot be written as a convex 
combination of other points in $\Conv(S)$. Importantly, every extreme point 
of $\Conv(S)$ lies in $S$. So, in Example~\ref{ex:simplex}, every 
$(S,k)$-sparse point is also $k$-sparse in the usual sense. In 
Example~\ref{ex:sdp}, every $(S,k)$-sparse point has rank at most $k$, since 
it is in the convex hull of $k$ rank-$1$ matrices.

\begin{example}
\label{ex:marginal-polytope}
Given a space $\sX$ and a function $\phi : \sX \to \bR^d$, let 
$S = \{\phi(x) \mid x \in \sX\}$. Then $\Conv(S)$ is called the 
\emph{marginal polytope} of $\sX$ with respect to $\phi$. The 
marginal polytope consists of all possible expected values of 
$\phi(x)$ with respect to an arbitrary probability distribution 
on $\sX$. A $(S,k)$-sparse point in $\Conv(S)$ is a point that 
can be written as a weighted average $\sum_{i=1}^k w_i \phi(x_i)$. 
In this case, we call the $x_i$ \emph{pseudosamples}.
\end{example}

\begin{example}
\label{ex:graph}
TODO: graph sparsification example
\end{example}

%% Many algorithms, such as SMC \citep{smc} and MCMC \citep{mcmc}, approximate a 
%% distribution with samples. Most such algorithms require access to a full posterior and 
%% may be prone to becoming stuck in local modes. In contrast, the herding algorithm introduced in 
%% \citet{Welling:2009a} generates samples whose moments are guaranteed to converge 
%% to any given collection of feasible moments. Moreover, \emph{only} the moments 
%% are required; a full posterior is not needed.
%% 
%% Since its introduction, herding has been shown to have many desirable properties.
%% For instance, \citet{Chen:2010a} showed that the moments converge at a rate of $O(1/T^{2})$ for 
%% $T$ samples, which is even faster than the $O(1/T)$ rate that would be obtained 
%% by sampling independently from the exact posterior. Furthermore, \citet{Bach:2012a} show that herding is 
%% equivalent to conditional gradient, and \citet{Huszar:2012} show that 
%% herding is equivalent to Bayesian quadrature. Herding was originally conceived of as a method 
%% for inference in Markov random fields, and this line of work has been further explored in 
%% \citet{Welling:2009a}, \citet{Gelfand:2010}, and \citet{Bornn:2013}.
%% 
%% In order to define herding, let $\sX$ be an observation space, $\phi : \sX \to \sF$ a 
%% feature map to an inner product feature space, and $\bar\phi = \bE_{x \sim p}[\phi(x)]$ the empirical 
%% moments of some distribution. Then the herding algorithm is the following simple procedure to 
%% generate pseudosamples from $\sX$:
%% \[
%% x_{t} &\in \argmin_{x \in \sX} \ip{\phi(x)}{y_{t-1}} \\
%% y_{t} &= y_{t-1}  + \bar \phi - \phi(x_{t}).
%% \]
%% In this paper we generalize the herding algorithm to minimize $\psi(x) + R^{*}(x)$ via the following set of updates, which 
%% require only the convexity of $\psi$ and $R$:
%% \[
%% x_t &\in \argmin_{x} \psi(x) + \ip{x}{y_{t-1}} \\
%% y_{t} &\in \argmax_{y} \ip{\bar{x}_t}{y} - R(y),
%% \]
%% where $\bar{x}_t$ is a certain weighted average of $x_1,\ldots,x_t$. Our 
%% algorithm is described in more detail in Section~\ref{sec:algorithm}. This 
%% generalization has several advantages:
%% \begin{itemize}
%% \item It allows us to cast herding in the framework of mirror descent \cite{Beck:2003}, 
%%       which lets us better understand its convergence properties.
%% \item It allows us to obtain generalizations of herding for infinite 
%%       dimensional spaces that apply in cases where standard herding does not.
%% \item It makes the known connections between herding, subgradient, and 
%%       conditional gradient methods more clear. 
%% \end{itemize}
%% In particular, while it was known that herding was a form of conditional 
%% gradient descent \cite{Bach:2012a} and hence (via the relationship in \citet{Bach:2012b}) also 
%% a form of subgradient descent, our framework exposes all three algorithms 
%% as special cases. A full table outlining the relationships for various choices of 
%% $\psi$ and $R$ is given in Table~\ref{tab:connections}.
%% 
%% In summary, we introduce a novel gradient descent algorithm to investigate herding and its generalizations, and make the following contributions:
%% \begin{itemize}
%% \item In Section \ref{sec:algorithm} we introduce a pair of convex optimization algorithms, which we call primal and dual boosted mirror descent (\primal and \dual). We show how \primal and \dual relate to conditional gradient (\cgd) and mirror descent (\md). 
%% \item We then use \dual in Section \ref{sec:herding} to define two classes of generalized herding algorithms. Connections with the ``learning'' interpretation of herding are made by showing how to obtain herding as the limit of an entropically normalized case of \dual. 
%% \item Section \ref{sec:proofs} presents convergence rates for \primal and \dual. As special cases, we recover a $O(\log T/ T)$ convergence rate for classical herding and a $O(1/T)$ rate for weighted herding. Both of these results apply to the infinite dimensional case. 
%% \item In Section \ref{sec:lower-bounds} we prove a matching lower bound to the upper bound given in Section \ref{sec:proofs}, showing that any ``sparse'' herding-like method cannot converge faster than $O(1/T)$ in the the infinite dimensional case.% and that herding converges at the slower $O(\log T / T)$ rate.
%% %\item Finally, in section \ref{sec:chen-proofs} we generalize the $O(1/T^{2})$ convergence proof of \citet{Chen:2010a} to a class of herding-like algorithms derived from \bmd. 
%% \end{itemize}
